---
title: "Cluster Testing: the two-tier model (Tier A on CRAN, Tier B on Hopper)"
vignette: >
  %\VignetteIndexEntry{Cluster Testing: the two-tier model (Tier A on CRAN, Tier B on Hopper)}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    code-fold: false
execute:
  eval: false
---

Every R chunk in this vignette is **shown, not run** (`execute: eval: false`).
The Tier-B suite it documents runs on a SLURM cluster, not inside
`R CMD check`, so building this vignette adds negligible time to a CRAN check
and never launches a cluster job. Outputs shown below are **static transcripts**
captured from a real Hopper run.

## Why two tiers

medsim's simulation and parallel code has two kinds of correctness property,
and they cannot both be checked in the same place:

| | **Tier A** | **Tier B** |
|---|---|---|
| Lives in | `tests/testthat/` | `inst/hopper-tests/` |
| Run by | `R CMD check`, `devtools::test()`, CI — **always** | manually on Hopper (`run-all.sh`) — **never by check** |
| Cores | 1 (`n_cores = 1`) | many (real FORK cluster) |
| Cost | seconds | a full SLURM array |
| Guards | pipeline correctness invariants: coverage discrimination, failure-rate accounting, truth-cache invalidation, chunk boundary cases | *at-scale* behaviour: many-core FORK RNG independence, production-grid seed collisions, full coverage at production reps |

`inst/` is installed with the package but **not executed** by `R CMD check`
(check runs only top-level `tests/*.R`), so the Tier-B scripts ship with the
package as runnable, documented artifacts yet cost the CRAN check nothing.

### The structural reason FORK-RNG can only be Tier B

The single most important guard — that parallel workers draw **independent**
random streams — cannot be tested in Tier A. Under `R CMD check`,
`_R_CHECK_LIMIT_CORES_=TRUE` is set, and medsim's `parallel.R` deliberately
forces sequential execution in that case:

```{r}
#| label: fork-limit
# R/parallel.R (paraphrased): under a CRAN check, never fan out.
if (isTRUE(as.logical(Sys.getenv("_R_CHECK_LIMIT_CORES_", "FALSE")))) {
  n_cores <- 1L   # forced sequential -> the FORK code path never executes
}
```

A Tier-A test therefore *cannot* exercise a real multi-core FORK cluster — the
one code path that the chunked-RNG bug lived in. That guard (B4 below) is
Tier-B by necessity, not by preference. Every other correctness invariant lives
in Tier A, where it runs on every commit.

## The self-contained synthetic study

Tier B does **not** depend on any downstream research kernel (e.g. the
product-of-three test-inversion CDF). It runs a self-contained
**Wald-CI-for-a-normal-mean** study: six scenarios with known true means, a
correct method, and a deliberately-broken control. This keeps the suite fast,
dependency-free, and independently checkable — the real research study stays
external as the genuine integration test.

```{r}
#| label: synthetic-study
# inst/hopper-tests/tier_b_synthetic.R (excerpt)
TIER_B_THETAS <- c(s1 = -1.0, s2 = -0.4, s3 = 0.0, s4 = 0.3, s5 = 1.0, s6 = 2.5)

# Correct method: a nominal 95% Wald interval for the mean.
tier_b_method_nominal <- function(data, params) {
  m  <- mean(data$x)
  se <- stats::sd(data$x) / sqrt(nrow(data))
  list(theta_lower = m - 1.96 * se, theta_upper = m + 1.96 * se)
}

# Planted defect: a 1/3-width interval that MUST undercover. This is the
# dogfood's can-fail control -- if coverage came out ~0.95 here, the whole
# harness would be silently broken.
tier_b_method_narrow <- function(data, params) {
  m  <- mean(data$x)
  se <- stats::sd(data$x) / sqrt(nrow(data))
  list(theta_lower = m - 1.96 * se / 3, theta_upper = m + 1.96 * se / 3)
}
```

## Running Tier B

On Hopper, after `R CMD INSTALL` of the current medsim, from
`inst/hopper-tests/`:

```{r}
#| label: run-all
# Submits PILOTS ONLY (<=4 array tasks each) -- never a full run.
# ./run-all.sh
```

`run-all.sh` submits small pilots and prints how to examine and combine them.
**Scaling to a full run is a deliberate, human-approved step** — see
"Pilot before scale" below.

### B2 — end-to-end coverage (the real number)

Runs `medsim_run_chunk` → `medsim_combine_chunks` → `medsim_analyze_coverage`
across a chunked array with the correct method; coverage must land near the
nominal 0.95.

```{r}
#| label: b2
# TIER_B_METHOD=nominal, then:
# Rscript combine_analyze.R tier_b_nominal nominal
```

Static transcript (full run, 20 chunks × 2000 reps):

```
=== Tier-B coverage (nominal), outdir=tier_b_nominal ===
  scenario parameter coverage n_valid n_failed failure_rate mean_width
1       s1     theta   0.9460    2000        0            0     0.2767
2       s2     theta   0.9455    2000        0            0     0.2769
3       s3     theta   0.9410    2000        0            0     0.2769
4       s4     theta   0.9500    2000        0            0     0.2769
5       s5     theta   0.9445    2000        0            0     0.2775
6       s6     theta   0.9525    2000        0            0     0.2768
overall mean coverage: 0.947  (n reps total: 12000)
verdict: PASS (near-nominal)
```

#### Combining runs under an integrity audit

`medsim_combine_chunks()` does more than concatenate: it rebuilds `$summary`
and `$config` over the combined frame and **audits the grid** — per-scenario
`replication` contiguity (ids are *global* across chunks under schema v2, so
gaps, duplicates, and ragged cells are all detectable), the seed-collapse
signature on every continuous estimate column, and cross-scenario seed
collisions. By default a violation **stops** the combine, signalling a
`medsim_combine_violation` condition that carries the combined object — no
cluster compute is lost:

```{r}
#| label: combine-audit
# Deliberate interim look at a still-running array (e.g. 58/60 chunks done):
partial <- medsim_combine_chunks("tier_b_nominal", on_violation = "warn")

# Recover the good cells from a stopped combine instead of losing the run:
combined <- tryCatch(
  medsim_combine_chunks("tier_b_nominal", expected_chunks = 20),
  medsim_combine_violation = function(e) e$results
)

# The same audit, standalone, on an already-combined object:
medsim_audit_results(combined, nsim = 2000)
```

### B3 — the can-fail dogfood (planted defect)

The same pipeline with the 1/3-width `narrow` method. If this *did not*
undercover, the harness could not detect a broken interval — so a pass here is
a pass of the **detector**, not just the method.

```{r}
#| label: b3
# TIER_B_METHOD=narrow, then:
# Rscript combine_analyze.R tier_b_narrow narrow
```

```
=== Tier-B coverage (narrow), outdir=tier_b_narrow ===
overall mean coverage: 0.480  (n reps total: 12000)
verdict: PASS (control undercovers as designed)
```

### B4 — many-core FORK RNG realism (the sole FORK guard)

A real FORK cluster over multiple cores: across the run, no two replications
may share a random draw, and re-running with the same seed must reproduce
exactly. This is the guard Tier A structurally cannot provide.

```{r}
#| label: b4
# Rscript fork_rng_realism.R   (TIER_B_N_REPLICATIONS controls the count)
```

```
reproducible (run1 == run2): TRUE
verdict: PASS
```

### B5 — production-grid seed-collision check

Verifies `.medsim_det_seed()` (the order-sensitive polynomial rolling hash that
fixed the chunked-RNG bug) produces **zero** collisions across the entire
production grid — not merely across the six sample scenario names.

```{r}
#| label: b5
# Rscript grid_collision_check.R tier_b_nominal
```

```
grid rows: 12000   distinct (scenario,CI): 12000
no cross-grid draw collision: TRUE
no .medsim_det_seed collision within any scenario: TRUE
verdict: PASS
```

## Pilot before scale

No full-scale cluster job is ever submitted cold. Every Tier-B check first runs
a **pilot** of ≤4 array tasks; the full run happens **only** after the pilot is
clean (all `COMPLETED`, `MaxRSS` within the `mem-per-cpu` envelope, nominal
coverage ~0.95, narrow control < 0.80, no collisions) **and** with explicit
approval. A failed or surprising pilot blocks the full run — fix and re-pilot.

```{r}
#| label: scale-check
# After the pilot: confirm State/MaxRSS/Elapsed before scaling up.
# sacct -j <jobid> --format=JobID,State,MaxRSS,Elapsed
```

The full run recorded above (jobs on UNM CARC Hopper, R 4.4.0) completed all 41
tasks in 2–4 s each with every check green. `MaxRSS` read below SLURM's ~30 s
sampler floor because the synthetic Wald workload is memory-trivial; a heavier
real-kernel study would register a meaningful `MaxRSS`, and the envelope should
be re-checked then.

## Summary

- **Tier A** (`tests/testthat/`, on CRAN) guards every pipeline correctness
  invariant that can run single-core in seconds.
- **Tier B** (`inst/hopper-tests/`, on Hopper) guards the at-scale and
  many-core-FORK properties that CRAN structurally cannot exercise.
- Building this vignette adds no measurable time to `R CMD check` and launches
  nothing — the cluster suite is documented here but run deliberately, by hand,
  behind a pilot gate.
